\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students

### 2023-06-15 phd-cs-us.tex
Updates from Overleaf

\documentclass[11pt]{article}
\usepackage[margin=1.1in]{geometry}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage[T1]{fontenc}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

@@

\newcommand{\mycomment}[3][\color{blue}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\didi}[1]{\mycomment{Didier}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen}\\
\href{https://cs.gmu.edu}{George Mason University}}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

\begin{abstract}
Having been involved in PhD admissions for many years, I've
realized that many international students, especially \emph{Vietnamese}, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to reveal some details about the PhD Admission process and share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is written for Vietnamese students interested in Computer Science, it should apply to students in many other countries and other majors.
Moreover, while many examples given are for George Mason University, the writing should generalize to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

In addition, this document can help \emph{US faculty} to understand more about international students, e.g., questions, concerns, and various cultural differences.  By understanding, appreciating, and even leveraging these information, CS programs in the US can attract larger and stronger application pools from Vietnam and other countries.

@@

\section{Should You Apply?}
\epigraph{Don't make fun of graduate students. They just made a terrible life choice.}{The Simpsons}

@@

Second, I believe that applying to a good US university \emph{should not} be any
harder than at schools in other countries. If you think you have a
chance at schools in other countries, e.g., South Korea, Singapore, Germany, UK, Japan and Australia, then you will surely have a chance in the US as well.

@@

\begin{commentbox}[Thanh]
Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have higly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs may present several challenges. Professors within these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, an 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. All of these make gaining admission in the US difficult as US faculty are not familiar with these issues.
\tcblower
\textbf{Vu}: Many Vietnamese students and faculty lament how this competitive/low-grading from top Vietnamese universities hurts its own students in applying abroad. One way to mitigate this is making these issues known to admission committee.  Schools with Vietnamese profs are probably aware of them, but in general your letter writers/you can explicitly mention these in their letters/ your statement.
\end{commentbox}

@@

Several about stipend. First, the amount of stipend depends on the university, which in turns depend on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is (typically) 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like for most source of income in the US, you will have to pay tax on your stipend. Finally, private universities might pay more for stipend.

\begin{commentbox}[Vu]
TA and RA at GMU have similar benefits in tuition waiving and insurance.  For stipend, depending on the college and department, a 9-month graduate assistant stipend is set.  TA and RA will usually be that amount (TA will definitely be that, RA might fluctuate a bit depending on the stage of the student and the prof.). 
\tcblower
Having health insurance is required at many US universities.  So do not think that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

TA is common in the beginning when you haven't found your advisor who would pay you RA. As a TA, you help professors with their classes (e.g., grading or teaching labs/recitation). Your TAship is paid through the school/department, i.e., they hire you to help teach.  During a semester, a TA might work with several courses and professors (not necessary their advisor).  TA funding typically is not available during the summer, which has no school.

@@

Even if you have other funding and do not need TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for research career where you often have to give talks and tell people about your work. Note that GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as a lecturer, which is higher than GTA.  This is a good opportunity for students to get teaching experience and also get paid more.
\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer. \S\ref{sec:ra-cost} gives more details on RA budget.

@@

\subsection{Tenured or tenure-track faculty? Who do you choose?} 
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.
Tenured faculty (e.g., associate and full profs.) are more likely to be older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured) and expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research, e.g., they are more involved in administrative duties.

>>>>

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[margin=1.0in]{geometry}
\usepackage{soul}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}

\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

@@

\newcommand{\mycomment}[3][\color{blue}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\didi}[1]{\mycomment{Didier}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen} (\href{mailto:tvn@gmu.edu}{tvn@gmu.edu})\\
 \href{https://cs.gmu.edu}{George Mason University}}

\begin{document}
\maketitle

\begin{abstract}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students}, especially those from smaller countries such as \emph{Vietnamese}, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to give some details about the PhD Admission process and share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is written for Vietnamese students interested in Computer Science, it should apply to students in many other countries and other majors.
Moreover, while many examples given are for George Mason University, the writing should generalize to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

In addition, this document can help \emph{US faculty and admission committee members} understand more about international students, e.g., questions, concerns, and various cultural differences.  By understanding, appreciating, and even leveraging these information, CS programs in the US can attract larger and stronger application pools from Vietnam and other countries.

@@

\section{Should You Apply?}
\epigraph{\vspace{-0.2in} Don't make fun of graduate students. They just made a terrible life choice.}{The Simpsons}

@@

Second, I believe that applying to a good US university \emph{should not} be any
harder than at schools in other countries. If you think you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan and Australia, then you will surely have a chance in the US as well.

@@

\begin{commentbox}[Thanh]
Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, an 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
\tcblower
\textbf{Vu}: Many Vietnamese students and faculty lament how this competitive/low-grading from top Vietnamese universities hurts its own students in applying abroad. One way to mitigate this is making these issues known to admission committee.  Schools with Vietnamese profs are probably aware of them, but in general your letter writers/you can explicitly mention these in their letters/ your statement.
\end{commentbox}

@@

Several words about stipend. First, the amount of stipend depends on the university, which in turns depend on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is (typically) 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like for most source of income in the US, you will have to pay tax on your stipend. Finally, private universities might pay more for stipend.

\begin{commentbox}[Vu]
TA and RA at GMU have similar benefits in tuition waiving and insurance.  For stipend, the college and department will set a 9-month graduate assistant stipend.  TA and RA will usually be that amount (TA will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof.). 
\tcblower
Having health insurance is required at many US universities.  So do not think that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

TA is common in the beginning when you haven't found your advisor who would pay you RA. As a TA, you spend up to 20 hrs/week and help professors with their classes (e.g., grading or teaching labs/recitation). Your TA is paid through the school/department, i.e., they hire you to help teach.  During a semester, a TA might work with several courses and professors (not necessary their advisor).  TA funding typically is not available during the summer, which has no school.

@@

Even if you have other funding and do not need TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for research career where you often give talks and tell people about your work. Note that GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as a lecturer, which is higher than GTA.  This is a good opportunity for students to get teaching experience and also get paid more.
\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer. \S\ref{sec:ra-cost} gives more details on RA budget.

@@

\subsection{Tenured or tenure-track faculty? Who do you choose?}

\begin{wrapfigure}{r}{0.48\textwidth}
    \vspace{-0.3in}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.
Tenured faculty (e.g., associate and full profs.) are more likely to be older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured) and expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research, e.g., they are more involved in administrative duties.
